---
title: GSE166927
output: html_notebook
---

## system

-   cmd+D: 删除光标所在当前行:

-   alt+-: 插入赋值号 \<- : Alt + -（R语言里几乎最常用的快捷键了）

-   插入管道操作符 : Ctrl + Shift + M（和tidyverse配套使用有奇效）

-   ctrl+shift+c: 注视掉所选代码块

-   执行行：cmd+enter

-   执行块：shift+cmd+enter

```{r}
system("open .")
```

```{r}
# BiocManager::install("BiocStyle")
# BiocManager::install("limma")
# BiocManager::install("Glimma")
# BiocManager::install("tidyverse",force = TRUE)
# BiocManager::install("Mus.musculus",force = TRUE)
# BiocManager::install("Homo.sapiens",force = TRUE)

# BiocManager::install("dplyr",force = TRUE)
# BiocManager::install("GEOquery",force = TRUE)
# BiocManager::install("tidyverse")
# BiocManager::install("Homo.sapiens")
```

## setup

```{r}
#setwd("/Users/m1/JobSkills/R/DEG/GSE117993/") 
```

```{r}
library(BiocStyle)
library(knitr)
options(digits=3)
options(width=120)
```

## library

```{r}
# library(tidyverse)
library(limma)
library(Glimma)
library(edgeR)
# library(tximport)
library(GEOquery)
library(Homo.sapiens)
library(recount3)
library(DESeq2)
```

-   [GSE117993](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE117993)

```{r}
library(tidyverse)
```

```{r}
rse <- recount3::create_rse_manual(
    project = "SRP155976",
    project_home = "data_sources/sra",
    organism = "human",
    annotation = "gencode_v26",
    type = "gene"
)
```

## read counts from recount

```{r}
assays(rse)$counts<-transform_counts(rse)
rse
# recount3::transform_counts()
```

获取group信息，分组信息，做差异表达必须。

```{r}
write.table(rse@colData, file = "colData.tsv", sep = "\t", row.names = FALSE, col.names = TRUE)
sample_title<-rse@colData$sra.sample_title
### 修剪字符串
sample_title<-sapply(sample_title, function(x) strsplit(x, split = "_")[[1]][4])
sample_title[1:10]
group <- sample_title
```

```{r}
unique(group)
```

## Gene ENSEMBL SYMBOL ENTREZID

```{r}
### convert exp matrix by ncbi gene, not ENSEMBL
ensg <- rownames(rse@assays@data$counts) %>% 
  sapply(function(x) substr(x, 1, 15))
### 注意这里因为冲突，所有要使用源代码的全名
genes <- AnnotationDbi::select(
  Homo.sapiens, 
  keys=ensg, 
  columns=c("SYMBOL", 
            "TXCHROM",
            "ENTREZID","ENSEMBL"), keytype="ENSEMBL"
  )
print("genes dim")
dim(genes)
# 75572     4

# genes<- genes %>% 
#   filter(!is.na(ENTREZID))
# dim(genes)

genes <- genes[!duplicated(genes$ENSEMBL),]
print("unique genes")
dim(genes)
# 63811     4
```

## DGEList

```{r}
x = new("DGEList")

x$counts = rse@assays@data$counts

x$samples = data.frame(
    "SampleID"=colnames(x$counts), 
     "group"=group,
    "lib.size"=colSums(x$counts),
    "norm.factors" = rep(1,190)
)

table(rowSums(x$counts==0)==190)
# FALSE  TRUE 
# 52486 11370 
table(rowSums(x$counts>=10)==190)
# FALSE  TRUE 
# 47242 16614 
print('full x')
dim(x)

cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE, prior.count=2)
dim(lcpm)

group<-x$samples$group
keep.exprs <- filterByExpr(x, min.count=10,group = group)
x <- x[keep.exprs, keep.lib.sizes=FALSE]
print('filterByExpr x')
# [1] 27806   190
dim(x)
```

add gene information to x

```{r}
x$genes <- genes
dim(x)
```

```{r}
# length(rownames(head(x$counts)))
dim(x$counts)
dim(x$genes)
    # x$genes$ENTREZID
```

## DESEQ2

```{r}
conditions <- factor(
    x$samples$group,
    levels = c("cCD","iCD","NotIBD","UC"))

colData <- data.frame(
    row.names = colnames(x$counts),
    condition=conditions)

dds <- DESeqDataSetFromMatrix(
    x$counts, 
    colData, 
    design = ~condition)

dim(dds@assays@data$counts)
# [1] 27806   190
dds <- DESeq(dds)
#获取结果
```

### p-value hist and output result

```{r}
results <- list()
for (t in c("cCD","iCD","UC")){
    print(t)
    results[[t]] <- as_tibble(results(dds, contrast = c("condition", "NotIBD", t)),rownames=NA)
    ### 补充基因信息
    results[[t]] <- results[[t]] %>%
      rownames_to_column(var = "RowNames") %>%
      mutate(ENSEMBL=substr(RowNames, 1, 15)) ### 修建ENSEMBL ID
    
   results[[t]] <- results[[t]] %>%
     left_join(genes[!duplicated(genes[["ENSEMBL"]]),], by = "ENSEMBL")
    write.table(results[[t]], file = paste("dds",t,'.tsv',sep="_"), sep = "\t", row.names = FALSE, col.names = TRUE)
    hist(results[[t]]$pvalue,breaks = 50)# ,xlim=c(0,1),ylim=c(0,2000)
}
```

### tydeverse

```{r}
dim(genes)
dim(filter(genes,!is.na(ENTREZID)))
```

```{r}
x$samples$lib.size
#[1] 32863052 35335491 57160817 51368625 75795034 60517657 55086324 21311068 19958838
#[1] 12854238 14470865 16528678 14493036 12386107 16700970 15207751 14601062 11996686 16650645 #12946464 13241413  9567869  9960254 12311350 10741452
#[17] 13193089 15132729 14964557  9337174 12133979 11472487 13432666 11887720
```

```{r}

L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
# summary(lcpm)
```

## filter

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
# -2.07
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,3))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)

for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", x$samples$SampleID, text.col=col, bty="n")
print('original')
dim(lcpm)

lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data default", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", x$samples$SampleID, text.col=col, bty="n")
print('filter')
dim(lcpm)

### 过滤 logCPM>=10的
keep <- (rowSums(lcpm<11) == 190) & (filterByExpr(x, min.count=20,group = group))
x <- x[keep, , keep.lib.sizes=FALSE]

lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. lcpm Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)

for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", x$samples$SampleID, text.col=col, bty="n")
print('filter by lcpm')
dim(lcpm)


```

## Filter DESeq2

```{r}
conditions <- factor(
    x$samples$group,
    levels = c("cCD","iCD","NotIBD","UC"))

colData <- data.frame(
    row.names = colnames(x$counts),
    condition=conditions)

dds <- DESeqDataSetFromMatrix(
    x$counts, 
    colData, 
    design = ~condition)

dim(dds@assays@data$counts)
# [1] 24557   190

dds <- DESeq(dds)
#获取结果

results <- list()
for (t in c("cCD","iCD","UC")){
    print(t)
    results[[t]] <- results(dds, contrast = c("condition", "NotIBD", t))
    hist(results[[t]]$pvalue,breaks = 50)# ,xlim=c(0,1),ylim=c(0,2000)
    write.table(results[[t]], file = paste("dds",t,'.tsv',sep="_"), sep = "\t", row.names = TRUE, col.names = TRUE)
}
# UC <- results(dds, contrast = c("condition", "NotIBD", "UC"))
# hist(UC$pvalue,breaks = 50)# ,xlim=c(0,1),ylim=c(0,2000)
```

```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors

```

## Normalization

```{r}
x2 <- x

par(mfrow=c(1,1))
lcpm <- cpm(x, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")
```

```{r}
x2 <- calcNormFactors(x2)
x2$samples$norm.factors
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")

```

```{r}
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- x$samples$group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- x$samples$group
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
plotMDS(lcpm, labels=x$samples$group, col=col.group)
title(main="A. Sample groups")
plotMDS(lcpm, labels=col.lane, col=col.lane, dim=c(3,4))
title(main="B. Sequencing lanes")

glMDSPlot(lcpm, labels=x$samples$group, 
          groups=x$samples[,c(2)], launch=FALSE)
```

## contr matrix

```{r}
group<-x$samples$group

design <- model.matrix(~0+group)
colnames(design) <- gsub("group", "", colnames(design))
design

contr.matrix <- makeContrasts(
   NotIBDvsUC        = NotIBD-UC,
   NotIBDvsiCD.      = NotIBD-iCD,
   NotIBDvscCD.      = NotIBD-cCD,
   levels = colnames(design))
contr.matrix

```

```{r}
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v

```

```{r}
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")

```

```{r}
## decidetests
summary(decideTests(efit))

tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)
```

```{r}
de.common <- which(dt[,1]!=0 & dt[,2]!=0)
length(de.common)
head(tfit$genes$SYMBOL[de.common], n=20)
vennDiagram(dt[,1:2], circle.col=c("turquoise", "salmon"))
write.fit(tfit, dt, file="results.txt")

```

```{r}
tfit
```

## p-value of EdgeR

```{r}
NotIBDvsUC <- topTreat(tfit, coef=1, n=Inf)
NotIBDvsiCD <- topTreat(tfit, coef=2, n=Inf)
NotIBDvscCD <- topTreat(tfit, coef=3, n=Inf)
par(mfrow=c(1,3))

hist(NotIBDvsUC$P.Value,breaks = 50,xlim=c(0,1),ylim=c(0,2000))
hist(NotIBDvsiCD$P.Value,breaks = 50,xlim=c(0,1),ylim=c(0,2000))
hist(NotIBDvscCD$P.Value,breaks = 50,xlim=c(0,1),ylim=c(0,2000))

```
