---
title: "R Notebook"
output: html_notebook
---

## library

```{r}
# library("tidyverse")
# library("recount3")
# library("dplyr")
# library("DESeq2")
# library("Homo.sapiens")
# # library("purrr")
# library("edgeR")
# BiocManager::install("GEOquery")
# BiocManager::install("limma")å
# BiocManager::install("umap")
# BiocManager::install("affy")
# BiocManager::install("annotate")
# BiocManager::install("hgu133plus2.db")


library(GEOquery)
library(limma)
library(umap)

library(dplyr)

library(affy)
library(annotate)
library(hgu133plus2.db)
```

```{r}

my_list <- list("apple", "banana", "cherry", "date")
positions <- c(4,1, 3,2)
selected <- my_list[positions]
 test_df <- data.frame(
   positions=positions
 )
 test_df$my_list <- selected
```

## code

```{r}
# Version info: R 4.2.2, Biobase 2.58.0, GEOquery 2.66.0, limma 3.54.0
################################################################
#   Data plots for selected GEO samples

# load series and platform data from GEO

gset <- getGEO("GSE30529", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL571", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

ex <- exprs(gset)
write.csv(ex,file = './ex.csv')

```

```{r}

gene_exprs <- exprs_annot %>%
  group_by(GeneSymbol) %>%
  summarise(across(where(is.numeric), max, na.rm=TRUE)) %>%
  column_to_rownames("GeneSymbol")


library(affy)
library(annotate)
library(hgu133plus2.db)

# eset 是 ExpressionSet 对象
probe_ids <- featureNames(eset)
gene_symbols <- getSYMBOL(probe_ids, "hgu133plus2")

exprs_df <- as.data.frame(exprs(eset))
exprs_df$GeneSymbol <- gene_symbols
```

```{r}

# log2 transform
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
  ex <- log2(ex) }

# box-and-whisker plot
par(mar=c(7,4,2,1))
title <- paste ("GSE30529", "/", annotation(gset), sep ="")
boxplot(ex, boxwex=0.7, notch=T, main=title, outline=FALSE, las=2)

# expression value distribution plot
par(mar=c(4,4,2,1))
title <- paste ("GSE30529", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, main=title, legend=F)

# mean-variance trend
ex <- na.omit(ex) # eliminate rows with NAs
plotSA(lmFit(ex), main="Mean variance trend, GSE30529")

# UMAP plot (multi-dimensional scaling)
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 9, random_state = 123)
plot(ump$layout, main="UMAP plot, nbrs=9", xlab="", ylab="", pch=20, cex=1.5)
# library("maptools")  # point labels without overlaps
# pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)
# 

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
